Simple repression draft

### 2020-05-28 doc/appendix_01_bayesian_inference.tex
MRM starting the intro section to the Bayesian inference appendix

% !TEX root = ./busty_transcription.tex
\section{Bayesian inference}
\label{sec:bayesian}
\mmnote{
\begin{itemize}
\item Bayes theorem.
Elaborate on posterior as plausibility over
space of model parameters.
\item Choosing likelihoods.
Easy here, just the steady-state distributions of our master
equations, assuming iid single-cell measurements.
\item Choosing priors.
Not critical here because we are data-rich and nonhierarchical,
as long as we don't exclude ground truth. In other applications,
prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling
assumptions and you just didn't realize it. Predictive checks to
verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}

\subsection{Parameter inference for constitutive promoters}

>>>>

% !TEX root = ./busty_transcription.tex
\section{Bayesian inference}
\label{sec:bayesian}

\subsection{The problem of parameter inference}

\mmnote{Bayes theorem. Elaborate on posterior as plausibility over space of
model parameters.}

One could argue that the whole goal of formulating theoretical models about
nature is to sharpen our understanding from qualitative statements to precise
quantitative assertions about the relevant features of the natural phenomena in
question \cite{Gunawardena2014}. It is in these models that we intend to distill
the essential parts of the object of study. Writing down such models leads to a
propagation of mathematical variables that parametrize our models. By assigning
numerical values to these parameters we can compute concrete predictions that
can be contrasted with experimental data. But for these predictions to match the
data the parameter values have to carefully be chosen from the whole parameter
space. But how do we go about asserting the effectiveness of different regions
of parameter space to speak to the ability of our model to reproduce the
experimental observations? The language of probability, and more specifically of
Bayesian statistics is -- we think -- the natural language to tackle this
question.

\subsubsection{Bayes theorem}

Bayes theorem is a simple mathematical statement that can apply to \textit{any}
logical conjecture. For two particular events $A$ and $B$ that potentially 
depend on each other Bayes theorem gives us a recipe for how to update our 
beliefs about one, let's say $B$, given that we get to observe something about
$A$. In its most classic form Bayes theorem is written as
\begin{equation}
P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
\end{equation}
where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
logical assertion. In particular the problem of Bayesian inference focuses on
the question of finding the probability distribution of a particular parameter
value given the data.

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given that we get to observe some
particular dataset. In other words, through the application of Bayes formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is caputerd in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

\subsubsection{The likelihood function}

\mmnote{Choosing likelihoods. Easy here, just the steady-state distributions of
our master equations, assuming iid single-cell measurements.}

As we alluded in the previous section it is through the likelihood function 
$P(D \mid \vec{\theta})$ that we encode the connection between our parameter 
values and the experimental observables. Broadly speaking there are two classes
of models that we might need to encode into our likelihood function:
\begin{itemize}
        \item Deterministic models: Models for which a concrete selection of
        parameter values give a single output. Say differently, rigid models 
        with a one-to-one mapping between inputs and outputs.
        \item Probabilistic models: As the name says it, models that rather than
        having a one-to-one mapping, describe the full probability distribution
        of possible outputs.
\end{itemize}
In this paper we focus on inference done with probabilistic models. After all
the chemical master equations we wrote down describe the time evolutions of the
mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will 
be given by the steady-state solution of the corresponding chemical master 
equation in question. This is rather convenient as we do not have to worry about
adding a statistical model on top of our model to describe deviations from the
predictions; our models themselves focus on predicting such variations from 
cell count to cell count.

\subsubsection{Prior selection}

\mmnote{Choosing priors. Not critical here because we are data-rich and
nonhierarchical, as long as we don't exclude ground truth. In other
applications, prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling assumptions and you
just didn't realize it. Predictive checks to verify you're not excluding any
plausible values.}

The different models explored in this work embraced different levels of 
coarse-graining that resulted in a diverse set of number of parameters. For each
of these model configurations Bayes theorem demands from us to represent our 
preconceptions on the possible parameter values in the form of the prior 
$P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we 
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is common practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomena in the form of a mathematical model -- which itself is a bold claim of
deeply understanding the features that matter in the object of study -- but we
have absolutely no idea what the parameter values ought to be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down prior distributions for our parameters.

\mmnote{
\begin{itemize}
% \item Bayes theorem.
% Elaborate on posterior as plausibility over
% space of model parameters.
% \item Choosing likelihoods.
% Easy here, just the steady-state distributions of our master
% equations, assuming iid single-cell measurements.
% \item Choosing priors.
% Not critical here because we are data-rich and nonhierarchical,
% as long as we don't exclude ground truth. In other applications,
% prior can matter a lot. This is a feature, not a bug: it's
% revealing hidden structure that was implied by your modeling
% assumptions and you just didn't realize it. Predictive checks to
% verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}

\subsection{Bayesian inference on constitutive promoters}
\label{sec:si_bayes_unreg}
